Supplmentary Material for DOI:

### 2023-07-11 manuscript/manuscript.qmd
biLSTM performance as two binary.

---
title: Estimating Sleep Quality Metrics Using Free-Living Accelerometer Data From Thigh-Worn Devices in Comparison to an EEG-Based Sleep Tracking Device
author:
  - name: Esben Høegholm Lykke
    email: eskovgaard@health.sdu.dk
    affiliations: 
        - id: some-tech
          name: University of Southern Denmark
          department: Department of Sports Science and Clinical Biomechanics
          address: Campusvej 55
          city: Odense
          state: Denmark
          postal-code: 5230
    attributes:
        corresponding: true
  - name: Jan Christian Brønd
    email: jbrond@health.sdu.dk
    affiliations:
        - id: another-u
          name: University of Southern Denmark
          department: Department of Sports Science and Clinical Biomechanics
          address: Campusvej 55
          city: Odense
          state: Denmark
          postal-code: 5230
abstract: "LÆS IKKE! Det er volapyk på latin. Lorem ipsum dolor sit amet, consectetur adipiscing elit. Suspendisse vitae dictum eros, ullamcorper elementum orci. Sed laoreet nulla neque, pulvinar fermentum mi iaculis at. Nulla ultricies nibh sit amet vestibulum rutrum. Nam pharetra nisl sed ipsum maximus suscipit. Duis metus nunc, ullamcorper eu mi rutrum, tempus ultricies ante. Nunc vitae lectus nisi. Aliquam efficitur ut eros ut pellentesque. Aenean blandit, nisl nec efficitur interdum, nisi ipsum fermentum dolor, at tempus sem turpis in lacus. Curabitur sollicitudin lectus sit amet velit pellentesque laoreet. Ut posuere diam lobortis nisi eleifend tincidunt. Ut at euismod sem, sed dignissim ligula. Aliquam lacinia massa libero, id eleifend velit pulvinar ac. Fusce volutpat elit eu nulla viverra, nec tempus orci pellentesque."
keywords: 
  - Sleep
  - Accelerometry
  - EEG
  - Machine learning
  - Sleep quality
date: last-modified
bibliography: bibliography.bib
format:
  elsevier-pdf:
    keep-tex: true
    journal:
      name: npj Digital Medicine
      formatting: preprint
      model: 3p
      cite-style: super
    include-in-header: 
      text: '\usepackage{lineno}\linenumbers'
    fig-pos: 'b'
  docx:
    number-sections: true
editor: 
  markdown: 
    wrap: 72
editor_options: 
  chunk_output_type: inline
---

@@

In this study, we employed a biLSTM, a multiclass classifier, to predict
three distinct states: out-of-bed-awake, in-bed-awake, and
in-bed-asleep. The architecture of the biLSTM was set up with four
layers, each equipped with 128 hidden units. This configuration was
intentionally chosen to balance between model complexity and training
efficiency: it provided the depth necessary for learning intricate
patterns while remaining feasible for timely training. The bidirectional
design of the LSTM served to enhance data interpretation and mitigate
overfitting by doubling the hidden units at each time step. For input,
we used tensors shaped as sequences, with each sequence spanning 10
minutes and a step size of one.This approach follows in the footsteps of
previous studies that utilized LSTM models for sleep detection. These
studies showcased the promising potential of LSTMs in capturing complex
temporal patterns. Particularly, the works of Sano et al. [@sano2019]
and Chen et al. [@chen2021] demonstrated the effectiveness of LSTM
models in improving sleep detection using accelerometer data,
underscoring the value of this modeling approach.

@@

In the context of our sequential learning strategy, the initial models
were tasked with the binary classification of in-bed vs. out-of-bed. For
this task, we assessed performance using the F1-score, accuracy,
sensitivity, specificity, and precision metrics. The second models in
our sequential learning strategy focused on the binary classification of
asleep vs. awake. For these models, we considered the same metrics, in
addition to the negative predictive rate. The class imbalance in this
case led us to compute the F1 score as an unweighted macro-average.
Additionally, we evaluated the multiclass classifier, biLSTM, using
macro-averaged F1-score, sensitivity, and precision. To further
illustrate model performance, we provide confusion matrices for the full
dataset, encompassing both in-bed and out-of-bed data. These matrices
report relative counts, column percentages (the proportion of the true
class accurately predicted), and row percentages (the proportion of
predictions correctly classified). We considered both the
in-bed/out-of-bed and awake/asleep scoring tasks as binary
classification problems, designating in-bed and asleep as the positive
labels and out-of-bed and awake as the negative labels in accordance
with previous research[@hjorth2012; @kushida2001].

To assess the performance of our models in deriving sleep quality
parameters, we utilized Bland-Altman plots and Pearson correlations. The
Bland-Altman method was employed specifically to determine the level of
agreement between two measurement techniques. Considering our dataset
contained multiple observations per subject, we integrated a bootstrap
procedure to address this extra source of variability. We calculated the
mean difference (bias) and defined the limits of agreement as the mean
difference plus or minus 1.96 times the standard deviation of these
differences. To ensure our measurements were robust and accounted for
intra-subject variability, we estimated the 95% confidence intervals for
both the bias and the limits of agreement using a bias-corrected and
accelerated bootstrap method, utilizing 10,000 bootstrap replicates. The
sleep quality parameteres included the following:

@@

4.  Latency Until Persistent Sleep (LPS) - This metric represents the
    time it takes to transition from wakefulness to sustained sleep. It
    is calculated as the time from the beginning of the ZM recording
    until a period when 10 out of 12 minutes are scored as sleep.

@@

```{r}
#| echo: false
#| message: false
#| label: tbl-in_bed_performance
#| tbl-cap: In-Bed Performance Metrics

@@

@tbl-sleep_performance details the performance of the second sequential
models on raw and median-filtered (5 and 10 minute) ZM predictions.
These are the predictions of the sequential models' ability to predict
asleep/awake on only the extracted in-bed time. The F1 scores, which are
unweighted macro averages, for raw ZM predictions range from 71.05% to
76.18%. The models perform comparably, but the low specificity values
(62.84% to 70.93%) suggest difficulty in correctly classifying awake
epochs. Applying 5-minute median filtering improves the performance
metrics. The XGBoost model tops the charts with an F1 score of 79.22%
and NPV of 74.00%. However, specificity still remains low, with values
between 54.68% and 74.84% across all models. With 10-minute median
filtering, the metrics improve further. The XGBoost model still leads
with an F1 score of 80.87% and an NPV of 75.76%. But, specificity
remains low, ranging from 57.47% to 76.35% across all models.

```{r}
#| echo: false
#| message: false
#| tbl-colwidths: auto
#| label: tbl-sleep_performance
#| tbl-cap: Sleep Performance Metrics

@@

The complete set of confusion matrices generated from data both
containing the out-of-bed and in-bed time are presented in
@fig-conf_mat. The figure shows favorable epoch-to-epoch performance
across across all sequential models, however, it is evident that the
biLSTM is less successful in classifying the in-bed-awake class which
cannot be deduced from the confusion matrices from the sequential
models. **\[fejl i biLSTM confmat. Alle mats er ens\]**

![Confusion matrices for binary sleep prediction. The middle of each
tile is the normalized count (overall percentage) and, beneath it, the
count. The bottom number is the column percentage (target). At the right
side of each tile is the row percentage (prediction). i) decision tree,
ii) logistic regression, iii) feed-forward neural net, iv) xgboost, and
v) biLSTM.](visuals/all_conf_mats.pdf){#fig-conf_mat}

@tbl-biLSTM_performance presents the performance of the three-class
biLSTM multiclassifier on raw and median-filtered (5-minute and
10-minute) ZM predictions. Raw ZM predictions achieve F1 Scores ranging
from 71.36% to 76.04%, indicating overall good performance. Applying
5-minute median filtering improves the metrics further, resulting in F1
Scores ranging from 75.99% to 78.53%, demonstrating enhanced precision
and sensitivity. **\[Overvej flere metrics. Evt. per class... pt er det
lidt svært at sammenligne på tværs af lstm og andre modeller ud over
F1.\]** With 10-minute median filtering, F1 Scores range from 73.45% to
73.93%, maintaining good performance but with limited information on
Specificity. Further analysis is required to assess performance across
all classes.

```{r}
#| echo: false
#| message: false
#| tbl-colwidths: auto
#| label: tbl-biLSTM_performance
#| tbl-cap: Performance of the three-class biLSTM multicclassifier.
source("code/table_biLSTM_performance.R")

tbl_biLSTM_performance
```

@@

@tbl-ba_cor presents a comparative analysis of the included models used
to predict various sleep quality parameters (SPT, TST, SE, LPS, WASO)
using the 5-minute median filtered ZM predictions. To see the full table
including models developed from raw ZM predictions and 10-minute median
filtered ZM predictions, see **\[SUPP. MAT.\]**. In terms of bias, the
decision tree model consistently underestimated SPT, TST, and SE, and
overestimated LPS and WASO in comparison to ZM. The logistic regression
model had similar trends, with more pronounced underestimation in TST
and overestimation in LPS. The eed-forward neural network also exhibited
similar bias as the decision tree and the logistic regression models,
but with a higher overestimation in WASO. On the other hand, the XGboost
model showed least bias among all, especially in its 5-minute median
predictions. Considering LOA, the decision tree had higher variability
across different sleep quality parameters and filtering techniques,
particularly for LPS and WASO, which indicates lower agreement with ZM.
Other models had comparable LOA but with notable exceptions. For
example, TST LOA for the logistic regression model was particularly wide
in the 5-minute median predictions. Correlation-wise, the pearson
coefficient, revealed that the XGboost model consistently had the
highest correlation with ZM across all sleep qualityparameters and
filtering methods Notably, the XGboost's 5-minute median predictions
showed the strongest correlation (0.66) for TST among all models and
filtering techniques.

```{r}
#| echo: false
#| message: false
#| label: tbl-ba_cor
#| tbl-cap: Summary of Bias, Limits of Agreement (LOA), and Pearson Correlation for various Sleep Parameter Predictions (SPT, TST, SE, LPS, WASO) using different Machine Learning Models (Decision Tree, Logistic Regression, Feed-Forward Neural Net, XGBoost) with Raw ZM Predictions, 5-Min and 10-Min Median as predictors. Each value is provided with its 95% Confidence Interval (CI).

@@

@fig-xgb_ba_cor shows the agreement of the XGBoost model trained on the
5-minute median filtered ZM predictions in estimating sleep quality
parameters as evaluated 5-minute median-smoothed ZM derived sleep
quality parameter. As can be seen on from the Bland-Altman plots, a
large portion of the nights are in almost perfect agreement with the ZM
and also a portion that can be considered outliers. This is true for all
sleep quality parameters.

**\[Overvej at smide samtlige 75 plots i supp.mat.\]**

![Comparison of sleep quality parameters derived from the XGboost model
trained on the 5-minute smoothed ZM predictions. The left column
displays Bland-Altman plots. Dashed lines represent the bias (the
average difference between the two measurements) and limits of
agreement, with the 95% confidence intervals represented as the grayed
areas. The right column displays scatter plots of XGboost-derived vs
ZM-derived sleep quality parameters. The dashed line represents the
identity line, while the full-drawn line represents the best linear fit.
Pearson's correlations are annotated in the upper left
corner.](visuals/median_5_xgboost_ba_cor.pdf){#fig-xgb_ba_cor}

@@

The present study embarked on an in-depth exploration of the utility and
comparative performance of different machine learning models in the
analysis of pediatric sleep data. The results offer promising insights
into the role of machine learning in enhancing the precision of sleep
quality metrics.

@@

It is unclear whether Johansson et al.[@johansson_development_2023]\]
provide a method for detecting the in-bed time...

>>>>

---
title: "Improving Sleep Quality Estimation: A Comparative Study of Machine Learning and Deep Learning Techniques Utilizing Free-Living Accelerometer Data from Thigh-Worn Devices and EEG-Based Sleep Tracking"
author:
  - name: Esben Høegholm Lykke
    email: eskovgaard@health.sdu.dk
    affiliations: 
        - id: some-tech
          name: University of Southern Denmark
          department: Department of Sports Science and Clinical Biomechanics
          address: Campusvej 55
          city: Odense
          state: Denmark
          postal-code: 5230
    attributes:
        corresponding: true
  - name: Jan Christian Brønd
    email: jbrond@health.sdu.dk
    affiliations:
        - id: another-u
          name: University of Southern Denmark
          department: Department of Sports Science and Clinical Biomechanics
          address: Campusvej 55
          city: Odense
          state: Denmark
          postal-code: 5230
abstract: "LÆS IKKE! Det er volapyk på latin. Lorem ipsum dolor sit amet, consectetur adipiscing elit. Suspendisse vitae dictum eros, ullamcorper elementum orci. Sed laoreet nulla neque, pulvinar fermentum mi iaculis at. Nulla ultricies nibh sit amet vestibulum rutrum. Nam pharetra nisl sed ipsum maximus suscipit. Duis metus nunc, ullamcorper eu mi rutrum, tempus ultricies ante. Nunc vitae lectus nisi. Aliquam efficitur ut eros ut pellentesque. Aenean blandit, nisl nec efficitur interdum, nisi ipsum fermentum dolor, at tempus sem turpis in lacus. Curabitur sollicitudin lectus sit amet velit pellentesque laoreet. Ut posuere diam lobortis nisi eleifend tincidunt. Ut at euismod sem, sed dignissim ligula. Aliquam lacinia massa libero, id eleifend velit pulvinar ac. Fusce volutpat elit eu nulla viverra, nec tempus orci pellentesque."
keywords: 
  - Sleep
  - Accelerometry
  - EEG
  - Machine learning
  - Sleep quality
date: last-modified
bibliography: bibliography.bib
format:
  elsevier-pdf:
    keep-tex: false
    journal:
      name: npj Digital Medicine
      formatting: preprint
      model: 3p
      cite-style: super
    include-in-header: 
      text: '\usepackage{lineno}\linenumbers'
    fig-pos: 'b'
  docx:
    number-sections: true
editor: 
  markdown: 
    wrap: 72
editor_options: 
  chunk_output_type: inline
---

@@

We also employed a biLSTM, a multiclass classifier, to predict three
distinct states: out-of-bed-awake, in-bed-awake, and in-bed-asleep. The
architecture of the biLSTM was set up with four layers, each equipped
with 128 hidden units. This configuration was intentionally chosen to
balance between model complexity and training efficiency: it provided
the depth necessary for learning intricate patterns while remaining
feasible for timely training. The bidirectional design of the LSTM
served to enhance data interpretation and mitigate overfitting by
doubling the hidden units at each time step. For input, we used tensors
shaped as sequences, with each sequence spanning 10 minutes and a step
size of one.This approach follows in the footsteps of previous studies
that utilized LSTM models for sleep detection. These studies showcased
the promising potential of LSTMs in capturing complex temporal patterns.
Particularly, the works of Sano et al. [@sano2019] and Chen et al.
[@chen2021] demonstrated the effectiveness of LSTM models in improving
sleep detection using accelerometer data, underscoring the value of this
modeling approach.

@@

In the context of our sequential learning strategy, the initial models
were tasked with the binary classification of in-bed vs. out-of-bed. For
this task, we assessed performance using the F1-score, accuracy,
sensitivity, specificity, and precision metrics. The second models in
our sequential learning strategy focused on the binary classification of
asleep vs. awake. For these models, we considered the same metrics, in
addition to the negative predictive rate. The class imbalance in this
case led us to compute the F1 score as an unweighted macro-average.
Additionally, we evaluated the multiclass classifier, biLSTM, using the
same metrics. To do this, we considered the multiclass output as to
binary classifications, where the first was out-of-bed vs the rest and
the second binary classification as in-bed-awake vs in-bed-asleep. To
further illustrate model performance, we provide confusion matrices for
the full dataset, encompassing both in-bed and out-of-bed data. These
matrices report relative counts, column percentages (the proportion of
the true class accurately predicted), and row percentages (the
proportion of predictions correctly classified). We considered both the
in-bed/out-of-bed and awake/asleep scoring tasks as binary
classification problems, designating in-bed and asleep as the positive
labels and out-of-bed and awake as the negative labels in accordance
with previous research[@hjorth2012; @kushida2001].

To assess the performance of our models in deriving sleep quality
parameters, we utilized Bland-Altman plots and Pearson correlations. The
Bland-Altman method was employed specifically to determine the level of
agreement between two measurement techniques. Considering our dataset
contained multiple observations per subject, we integrated a bootstrap
procedure to address this extra source of variability. We calculated the
mean difference (bias) and defined the LOA as the mean difference plus
or minus 1.96 times the standard deviation of these differences. To
ensure our measurements were robust and accounted for intra-subject
variability, we estimated the 95% confidence intervals for both the bias
and the LOA using a bias-corrected and accelerated bootstrap method,
utilizing 10,000 bootstrap replicates. The sleep quality parameteres
included are defined as follews in accordance with the ZM definitions:

@@

4.  Latency Until Persistent Sleep (LPS) - This metric represents the
    time it takes to transition from wakefulness to sustained sleep. It
    is calculated as the time from the beginning of the ZM recording
    until the first period when 10 out of 12 minutes are scored as
    sleep.

@@

```{r}
#| echo: false
#| message: false
#| label: tbl-in_bed_performance
#| tbl-cap: In-bed performance metrics

@@

@tbl-sleep_performance details the performance of all sequential model
types on raw and median-filtered (5 and 10 minute) ZM predictions for
sleep/wake classification. For raw ZM predictions, the F1 scores, which
are unweighted macro averages, range from 65.6% (biLSTM) to 76.2%
(XGBoost). All models perform comparably, but the low specificity values
(62.5% to 70.9%) suggest difficulty in correctly classifying awake
epochs. Applying a 5-minute median filter improves the performance
metrics. The XGBoost model tops the charts with an F1 score of 79.2% and
NPV of 74.0%. However, specificity still remains low, with values
between 54.7% (XGBoost) and 74.8% (Logistic Regression) across all
models. With a 10-minute median filter, the metrics improve further. The
XGBoost model still leads with an F1 score of 80.9% and an NPV of 75.9%.
But, specificity remains low, ranging from 57.5% (Decision Tree) to
76.4% (Logistic Regression) across all models.

```{r}
#| echo: false
#| message: false
#| tbl-colwidths: auto
#| label: tbl-sleep_performance
#| tbl-cap: Performance of the sleep/wake classification of the sequential models.

@@

A complete set of confusion matrices generated from data both containing
the out-of-bed and in-bed time are presented in @fig-conf_mat. The
figure shows favorable epoch-to-epoch performance across across all
sequential models, however, it is evident that the biLSTM is less
successful in classifying the in-bed-awake class which cannot be deduced
from the confusion matrices from the sequential models.

![Confusion matrices for binary sleep prediction. The middle of each
tile is the normalized count (overall percentage) and, beneath it, the
count. The bottom number is the column percentage (target). At the right
side of each tile is the row percentage (prediction). i) decision tree,
ii) logistic regression, iii) feed-forward neural net, iv) XGBoost, and
v) biLSTM.](visuals/all_conf_mats.pdf){#fig-conf_mat}

@@

@tbl-ba_cor presents a comparative analysis of the included models used
to predict various sleep quality parameters (SPT, TST, SE, LPS, WASO)
using the 5-minute median filtered ZM predictions. To see the full table
including models developed from raw ZM predictions and 10-minute median
filtered ZM predictions, see **\[SUPP. MAT.\]**. In terms of bias, the
decision tree model consistently underestimated SPT, TST, and SE, and
overestimated LPS and WASO in comparison to ZM. The logistic regression
model had similar trends, with more pronounced underestimation in TST
and overestimation in LPS. The eed-forward neural network also exhibited
similar bias as the decision tree and the logistic regression models,
but with a higher overestimation in WASO. On the other hand, the XGBoost
model showed least bias among all, especially in its 5-minute median
predictions. Considering LOA, the decision tree had higher variability
across different sleep quality parameters and filtering techniques,
particularly for LPS and WASO, which indicates lower agreement with ZM.
Other models had comparable LOA but with notable exceptions. For
example, TST LOA for the logistic regression model was particularly wide
in the 5-minute median predictions. Correlation-wise, the pearson
coefficient, revealed that the XGBoost model consistently had the
highest correlation with ZM across all sleep qualityparameters and
filtering methods Notably, the XGBoost's 5-minute median predictions
showed the strongest correlation (0.66) for TST among all models and
filtering techniques.

```{r}
#| echo: false
#| message: false
#| label: tbl-ba_cor
#| tbl-cap: Summary of Bias, Limits of Agreement, and Pearson Correlation for various Sleep Parameter Predictions (SPT, TST, SE, LPS, WASO) using different Machine Learning Models (Decision Tree, Logistic Regression, Feed-Forward Neural Net, XGBoost) with Raw ZM Predictions, 5-Min and 10-Min Median as predictors. Each value is provided with its 95% Confidence Interval (CI).

@@

@fig-xgb_ba_cor shows the agreement between the XGBoost model, trained
on 5-minute median filtered ZM predictions, and the 5-minute
median-smoothed ZM-derived sleep quality parameters. The Bland-Altman
plot for the SPT reveals a significant level of agreement with the ZM,
as evidenced by a positive correlation. However, the presence of extreme
outliers widens the limits of agreement (LOA). The scatterplot for SPT
also demonstrates a positive trend, indicating a moderate linear
correlation between the XGBoost model and the ZM-derived sleep quality
parameters.In terms of TST, perfect agreement is not observed for a
substantial number of nights, but there is a positive correlation
between the XGBoost model and ZM-derived sleep quality parameters. The
bias and LOA for TST are comparable to those observed for SPT,
indicating a consistent level of agreement between the two methods. The
scatterplot for TST also shows a slightly higher correlation, primarily
driven by the absence of extreme outliers.Furthermore, the remaining
three sleep quality parameters, SE, LPS, and WASO, exhibit
heteroscedasticity in contrast to SPT and TST. This outcome is expected
as achieving 100% sleep efficiency is relatively rare, resulting in less
disagreement between the methods as values approach the upper limit.
However, as sleep efficiency decreases, the potential for discrepancies
and differing interpretations between the methods increases, leading to
greater heteroscedasticity. A moderate linear correlation is observed
between the XGBoost model and ZM-derived sleep quality parameters for
SE, indicating a positive relationship. However, a poor correlation is
observed for LPS and WASO, suggesting less agreement between the two
methods for these parameters. Similar plots for all models are available
in the supplementary materials.

![Comparison of sleep quality parameters derived from the XGBoost model
trained on the 5-minute smoothed ZM predictions. The left column
displays Bland-Altman plots. Dashed lines represent the bias (the
average difference between the two measurements) and LOA, with the 95%
confidence intervals represented as the grayed areas. The right column
displays scatter plots of XGBoost-derived vs ZM-derived sleep quality
parameters. The dashed line represents the identity line, while the
full-drawn line represents the best linear fit. Pearson's correlations
are annotated in the upper left
corner.](visuals/median_5_xgboost_ba_cor.pdf){#fig-xgb_ba_cor}

@@

In an effort to mature the methods for estimating sleep from thigh-worn
accelerometers, we evaluated various models for predicting in-bed and
sleep time and their derived sleep quality parameters. Furthermore, we
trained and evaluated the models using raw and median-filtered
gold-standard predictions from the ZM. In general, all sequential models
performed well at predicting in-bed time. More challenging was it to
distinguish wake from sleep on the extracted in-bed time, however, the
performance of the sequential models were enhanced by the application of
median filters. Moreover, even though the multiclass biLSTM showed good
performance across F1 score, precision and NPV, the derived sleep
quality parameters were not on par with the XGBoost model which
demonstrated the highest performance metrics across all evaluations,
including epoch-to-epoch prediction and sleep quality parameter
derivatives. Despite this, all sequential models showed low specificity
values, indicating difficulty in correctly classifying awake epochs
during time in bed. The application of 5-minute and 10-minute median
filters improved the performance metrics of all models, with the XGBoost
model consistently leading. The filters increased total sleep time and
sleep efficiency, while reducing wake after sleep onset and the number
of awakenings. The XGBoost model also showed the least bias and highest
correlation with ZM across sleep quality parameters.

Limited research exists regarding the epoch-to-epoch effectiveness of
classifying in-bed time based on data from thigh-worn accelerometers.
Nevertheless, Carlson and colleagues provided compelling insights. They
demonstrated that a third-party algorithm, "ProcessingPal," and a
proprietary one, "CREA," achieved accuracies of 91% and 86%
respectively. These algorithms, evaluated against self-reported
measures[@carlson2021], produced F1 scores as high as 95% and 96%. These
figures are consistent with the performance of our sequential models,
which also achieved F1 scores and accuracy scores exceeding 95% in
identifying in-bed time. In our study, in-bed time is equated with SPT.
All models, with the exception of XGBoost, underestimated SPT. The
biLSTM model showed the greatest underestimation, with a bias of -36
minutes, reflecting trends observed in previous research. Winkler et al.
developed an algorithm that, despite a strong correlation (Pearson
correlation coefficient = .67) between their algorithmic results and
diary-recorded waking times, overestimated waking wear time by more than
30 minutes, resulting in an underestimation of in-bed
time[@winkler2016]. This trend was further confirmed when Inan-Eroglu et
al. examined Winkler et al.'s algorithm, revealing a underestimation of
9.8 minutes in bed time compared to self-reported
measures[@inan-eroglu2021]. In contrast, a study by van der Berg et al.
reported a slight underestimation of in-bed time. They employed a unique
approach with their algorithm, which relied on quantifying the number
and duration of sedentary periods to determine time in bed, and active
periods (standing or stepping) to identify wake times[@vanderberg2016].
Finally. it is important to note that high predictive performance in
determining in-bed time does not necessarily translate to accurate
predictions of broader sleep quality parameters. The crucial task of
detecting awake periods during in-bed time, a key factor in assessing
sleep quality, may not be effectively captured by in-bed time
predictions alone. Indeed, underestimating in-bed time could result in
overestimating waking time during in-bed time. Furthermore, the
distinction between actual sleep and time spent in bed, often overlooked
but vital in sleep research, is critical for a comprehensive
understanding of sleep quality.

To the best of our knowledge, Johansson and
colleagues[@johansson_development_2023] are the only researchers who
have reported epoch-to-epoch performance metrics for sleep scoring using
thigh-worn accelerometers, beyond just "waking time" and "in-bed time."
They achieved a mean sensitivity of 0.84, specificity of 0.55, and
accuracy of 0.80, using a single-night evaluation dataset of 71
subjects. Despite our models achieving a sensitivity above 97%, they,
like Johansson et al.'s algorithm, struggled with detecting in-bed awake
epochs. This is reflected in the low specificity scores, ranging from
54.7% to 76.4%, reported in our study. The challenge of low specificity
is not unique to methods using thigh-worn devices. Conley et al.'s
meta-analysis[@conley2019] reported similar findings when estimating
sleep using wrist-worn accelerometers among healthy adults, with a mean
sensitivity, accuracy, and specificity of 0.89, 0.88, and 0.53,
respectively. These figures align with our results regarding in-bed
time. Patterson and colleagues[@patterson_40_2023] recently summarized
the performance of various heuristic algorithms, machine learning, and
deep learning models used to predict sleep. They found the mean
sensitivity and specificity to be 93% (SD = 2.8) and 60% (SD = 11.1)
respectively. These findings underscore the challenge of automating the
detection of in-bed awake periods. Interestingly, despite low
specificity values for most of our models and configurations, we
observed an overestimation of LPS and WASO, contrasting with most
previous research[@conley2019; @palotti2019]. This overestimation of
wake epochs is evident from the low NPV scores, indicating that only a
small proportion of the wake predictions are actually correct. This
discrepancy may be driven by the SMOTE process used to balance the
dataset. If the synthetic "wake" examples created by SMOTE are not
representative of the true "wake" data, the models might learn to
incorrectly classify certain "sleep" epochs as "wake". This could lead
to an overestimation of LPS and WASO, as the models are incorrectly
identifying more periods of wakefulness during the sleep period.

often times imbalanced datasets only containing night recordings

The sleep quality parameter that our models consistently corresponded
most with the ZM derived sleep quality parameters were SPT and TST

something on outliers...maybe related to incl 180 degrees problem.

single night vs multiple nights/days

It is unclear whether Johansson et al.[@johansson_development_2023]\]
provide a method for detecting the in-bed time which hinders its
applicability for researchers that want to ...

@@

strengths and limitations

ZM as gold-standard? Without a gold standard, both methods can both
contribute to disagreement.

external validation dataset/generalization

expand the feature set (Palotti)

### 2023-07-11 manuscript/supp_mat/supp.qmd
biLSTM performance as two binary.

![all ba and scatterplots](../visuals/all_ba_scatterplots.pdf)

>>>>

![Decision tree, raw](../visuals/raw_decision_tree_ba_cor.pdf){#tbl-raw_dc}
![Decision tree, median_5](../visuals/median_5_decision_tree_ba_cor.pdf){#tbl-m5_dc}
![Decision tree, median_10](../visuals/median_10_decision_tree_ba_cor.pdf){#tbl-m10_dc}

![Logistic regression, raw](../visuals/raw_logistic_regression_ba_cor.pdf)
![Logistic regression, median_5](../visuals/median_5_logistic_regression_ba_cor.pdf)
![Logistic regression, median_10](../visuals/median_10_logistic_regression_ba_cor.pdf)

![Neural_network, raw](../visuals/raw_neural_network_ba_cor.pdf)
![Neural_network, median_5](../visuals/median_5_neural_network_ba_cor.pdf)
![Neural_network, median_10](../visuals/median_10_neural_network_ba_cor.pdf)

![XGBoost, raw](../visuals/raw_xgboost_ba_cor.pdf)
![XGBoost, median_5](../visuals/median_5_xgboost_ba_cor.pdf)
![XGBoost, median_10](../visuals/median_10_xgboost_ba_cor.pdf)

![biLSTM, raw](../visuals/raw_biLSTM_ba_cor.pdf)
![biLSTM, median_5](../visuals/median_5_biLSTM_ba_cor.pdf)
![biLSTM, median_10](../visuals/median_10_biLSTM_ba_cor.pdf)
